CHALLENGES TO PROFESSIONAL FOOTBALL COMPANIES AND THEIR ANSWERS WITH PARTICULAR REGARD TO ORGANISATIONAL CHANGES

Éva Bába Ms Bács1[P1]

Abstract
Professional football has been going through a period of unprecedented economic growth since the ?90s. Football ventures are increasingly becoming medium-sized companies. There have also been organizational changes, reflected in changes in the legal forms of professional football clubs, and in the use of modern controlling, planning, risk and financial management. Important questions remain unanswered with regard to the financing of football clubs, such as the impact of risks or the market value of capital costs. In addition to liquidity and the above, the acceptance of the determinational dependence of certain capital funds on sports results and the development of a strategy suitable for the optimal target system are also important requirements. In this article, we would like to present the answers football companies have given to the challenges they have been facing which affect their organisational system. In the light of international comparison, we examine the status of Hungarian football, which was once world famous and enjoyed better times in the past.

Keywords
Professional Football, Legal Forms, Organizational Structure, Controlling, Controlling System Building in Football Undertakings

I. Introduction
Since the 1990s, professional football has undergone an unprecedented economic resurgence worldwide. From an economic point of view, this sport has evolved into an industry that goes beyond the football clubs participating in the championships that are organized by the various leagues to also involve rather untransparent, complementary economic operations where companies increasingly perform their specific activities on the level of medium-sized enterprises ? measured against traditional industrial actors ? by following their economic interests.
In line with these truly economic processes, football in Hungary should also take steps to become an industry and apply the methods of more developed markets to the work of the Hungarian Football Federation (MLSZ), as well as to its organizational setup and the management of the Football League and professional clubs, both in their acquisition of capital resources and the establishment of control (Borbély, 2012). This is the only way to ensure the long-term, efficient operation of Hungarian clubs in the international scene, the further betterment of the standards of the OTP Bank League, something which is key to attaining positive results for the Hungarian national team.
The general conditions of professional football change at such a pace that most football clubs find it hard to follow in terms of adaptiveness. These changes are so intense that one cannot avoid agreeing with Caesar Luis Menotti, the legendary Argentinean coach: "The world of utopias has now gone. We live in a utilitarian society where football is condemned to big businesses." 
II. Changes in professional football
Over the past few decades, tremendous changes have taken place in football, both on and off the pitch. We would like to highlight just a single piece of data to illustrate this: in 1963, the monthly salary was DM 1,200 in the level one Bundesliga in Germany. Only the world champion Hans Schäfer received double this. In 2005, the average monthly salary reached EUR 26,000-27,000, while the biggest stars were even then earning millions each year. The ?off the pitch? changes can be divided into four groups.

Figure 1: Categorisation of ?off the pitch? changes



Source: Own research

The above revolutionary changes are closely interconnected with each other.

Financial field
Sales revenues, especially in the case of the teams of big football nations, have increased dramatically in recent decades. The German level one Bundesliga had already reached EUR 2 billion in sales revenue as early as in 2010, while for England revenue exceeded EUR 2.4 billion. The numbers are immense, for example in 2010 in Germany, the average number of spectators visiting a match was about 41,000, with a standard ticket price of more than EUR 21. Media revenues have also massively increased. It is highly important to note that professional football is a very costly economic field, where an increase in sales revenues goes hand-in-hand with considerable growth in expenses. Financial stability plays a huge role, and this is reflected in the quality of the management and its setup. From the perspective of our selected topic, organisational changes in the latter and the establishment of controlling organisations have great significance (Haas, 2006).
Capital can be provided through internal or external financing. Football enterprises are highly dependent on internal financing, where forms of financing specific to football are especially prevalent. Financing sources in this case can be expanded by increasing revenues and decreasing expenditures or by releasing tied up capital. The main revenues in a classical sense are revenues from media rights and ticket sales, merchandising (image transfer), and revenues from noneconomic areas or areas close to football, which in the case of a football enterprise obviously depend on sports results (Keller, 2006).
In the case of significant economic and intangible results, the dependence of primary revenues on sports results significantly decreases. Examples of this are media rights revenues, which are sold through contracts prior to the launch of the season. For certain periods, these revenues can be considered fixed. In this case, growth opportunities are of course highly limited for the duration of these contracts. Revenue growth can only be achieved thanks to sports results, unexpected successes abroad or by the integrating of new media elements. Revenues from media rights can therefore be considered a given condition depending on the previous season, which of course does not mean that the club should not try to generate as much revenue as possible, as media revenues are usually the most significant and most intensely growing revenue source for professional football clubs. In Germany, in the 2008/2009 season in the Bundesliga I, such revenues increased by 40 percent compared to the 2006/2007 season (4. Bundesligareport, 2011). The revenue structure of the 2009/2010 season is shown in Table 1. The tendency observed is that clubs have to move in the direction of pay-per-view channels instead of free television channels. This is more apparent in Italy, France and England, where revenues are greater as a result. Significant growth is expected in the field of secondary sales and foreign broadcasts, where the main target countries are China and Japan. The role of new media (Internet and mobile) is also on the rise.

Table 1: German football championship revenues in the 2009/2010 season (Bundesliga 1), in billions of euro and %
Game and  ticket revenues0.37921.43%Transfer revenues0.1066.01%Sponsorships and advertising0.51228.92%Merchandising revenues0.0744.7%Media revenues0.50528.55%Other revenues0.19310.93%Total                                                                                                                             1,769   100%
Source: 4. Bundesligareport 2011, www.bundesliga.de

Legal field and international balance sheets
The legal form, structure, organisational setup and management of businesses considered medium-sized even in industrial terms have been necessarily adjusted to changes in the market and the environment. Companies have basically been formed in the following two ways:
a) The decision of the Football Association makes the outsourcing of the licence department and its transformation into a company possible (e.g. Germany) 
b) The Football Association makes the formation of a capital company a requirement in level one or in both top levels.

Result factors (sports, economic and intangible results) are interrelated and interdependent, with their sum giving us the aggregate result (Bott, 2007). Naturally, the exploitation of opportunities arising due to the operation of incorporated companies emphasises the maximisation of results within the operation. (It would be expedient to discuss legal forms separately, as their significance within football makes this necessary). The economic result of a football enterprise should never be identified as the profit achieved in the fiscal year, as the latter is no more than a snapshot of long-term economic performance. A more interesting issue is what economic activity was used to generate the economic profit of the given year; whether it is the result of the previous championship season, and whether, taking into account the given sports result risks, these economic conditions will be repeated or not in the next season.
The intangible result, the success of a football enterprise, entails value factors that are very difficult to quantify using financial-economic or sports result aspects. This includes the image of the football enterprise, the improvement of which usually results in an increased media presence, the improvement of financial positions or the growth of the market value of the enterprise. Of course this also impacts economic results, as an improved image could lead to increased sponsorial involvement. When speaking of intangible results, we should also mention the supporters? strong identification with the club. This strong relationship has a positive impact not only on the economic result, but could also boost the achievement of better sports results. (Many have experienced the legendary ?Fradi-heart?, and probably even more have heard of it). But it is important to encourage, reward and pay the players as well (Nagy and Való, 2013). The personal goals of competent club executives could also strongly influence the intangible results of football enterprises. In this case, professional decision-making considerations play only a subordinated role, and in most cases this particular factor is enforced through the sympathy or dislike shown towards club executives or owners or their acceptance and ability to enforce their interests (Borbély, 2012).
A football enterprise naturally strives to maximise its intangible results and value; however, this can only be achieved at the expense of other result factors, and primarily clashes with the economic result. If the club?s management and owner decide to sign valuable star players and managers regardless of the size of the expenditure, this will obviously have a positive effect on the intangible results and naturally on the sports results as well. It is very likely that in the long-term the enterprise?s revenues will also increase. In the mid-term, however, such decisions impose significant financial burdens, which could in turn lead to a serious financial crisis for the club. In professional football ? as opposed to other economic fields ? it is not customary to prepare a separate plan with respect to impact on image. Today, change in image is still an indirect effect of investment. The way of the future is that football enterprises shall directly plan to increase intangible results through certain targeted promotions, e.g. by planning and managing member recruitment, loyalty promotions, etc.
The three main components of the aggregate result and their factors are closely interrelated. For example, the improvement of sports results has a positive impact on the economic result, which in turn leads to an increase in television and ticket revenues and the sale of promotional products, but member contributions and revenues from sponsorship deals could also grow. Improving sports results also affect intangible results: the club becomes more respected, supporter ties become stronger and media interest increases. A good economic result has a favourable effect on sports results, as the sufficient liquidity provided through profit and appropriate management allows the club to provide an incentive wage and premium payments without any problems. We should also emphasise that a good economic result also has a positive effect on the intangible result. Opportunities open up to finance supporter projects and clubs as well as significant advertising activity. Star players and managers can be signed to the extent of expanding financial opportunities. The intangible result also has an effect on the factors of the other two result components: sports results improve through the impact on public opinion, but the economic result also improves, since television revenues and the sale of endorsed products increase. In the event of sufficient interest, live broadcasts can be set up, while member contributions as well as sponsorship revenues rise. Understandably, sponsors highly appreciate the significance of image (Allen, 2011).
A football enterprise acts correctly if it strives to achieve the maximum of the composite result made up of the aforementioned three components.

Marketing field
Presence in the media is of paramount importance. The football club is sold as a brand value. Diversification activities and globalisation efforts appear followed by the entry into the large Asian and other foreign markets with modern media tools.   

Environment
At modern and successful professional football teams, the number of stakeholders increases significantly and harmonising their various needs is a very high level management task.
III. Organizational adaptation of football enterprises
Football is now the scene of structural changes worldwide that are embodied in the alteration of the legal forms of enterprises backing professional football, as well as the spreading application of modern controlling, planning, risk and financial management tools. From an economic perspective, with these recently emerging elements, football clubs have started to build up a modern and complex, specialized sector where the most important requirement until the end of the championship season is the maintenance of licensing conditions.

Role and significance of legal forms and organizations of enterprises in football
In line with the international outlook, I primarily wish to rely on the peculiarities of German professional football, since it best represents the presumable mainstream of the development of football. Let us review the forms of undertaking in German League football, and what the changes in trends are.
On the basis of the 1998 Resolution of the German Federal Parliament, it is also possible to operate football undertakings in the form of an incorporated firm, which means that members are not responsible for the liabilities of the company (these are share companies, limited liability companies, and limited share partnerships), in addition to the classical form of association (official Bundesliga website, 2011). The main point here is that part of the association participating in League football can be outsourced into the form of incorporated firm. It is interesting that, in Germany, 21 teams from League 1 and 36 teams from League 2 are currently operating in the form of incorporated firms (Figure 2).

Figure 2: Distribution of legal status in Germany (I-II. League clubs)

Legal FormFrequencyAssociation15Share comp.2LLC4LLC with right5LLC & Co.LP.10I.-II Liga , Total36
Source: Dworak, 2010

Advantages and disadvantages of registered association
Funding, registering and operating an association is relatively simple, and the rules are transparent. All strata of society like to make use of the advantages of this form of a non-profit company having a legal personality which provides tax benefits. The registered associations fill such an important social role in Germany that they are exempted, in respect of public purpose activity, from corporate income tax, industrial tax, real estate and property tax, and in addition are only charged a beneficial 7% general turnover tax (VAT). Normally, the general rate of VAT is 19%. However, these tax benefits affect professional football undertakings to only a minor extent, as these professional associations may obtain only a small amount of revenues exempt from taxation. The tax benefit mostly helps amateur and mass sports. Unfortunately, significant is the tax evasion (Nagy, 2011). A further drawback of operations in the form of associations is that there are no rules concerning the distribution and utilization of the realized profit. In consequence, most of the football clubs expend realized profits on efforts to increase the value of the team of players, while having a much weaker focus on raising financial reserves, which would otherwise be beneficial for long-term operations.
The rules relating to associations do not provide satisfactory protection even for lenders. Associations are answerable with their own assets for their liabilities, but there are no detailed legal prescriptions in the relevant law of associations in Germany. Of course, in principle, regulations may be set up relating to publicity, application of profits and collateralising loans in the statutes of associations, which is, however, scarcely feasible. We can hardly suppose that the elected management of an association would set up restrictions on itself. Logically, all this has a negative impact on the opportunities for financing with outside capital.
It follows from all this that a registered association may only make restricted use of any potential financing options. An increase in equity is allowed primarily through the retention of profits, but we must add that the profits may only be increased up to a certain limit per year, in order to maintain the non-profit status of the company. The hosting of shares and the involvement of contributed capital may only be secured after transformation into an incorporated firm. Therefore, in the event of associations, funding by banks or private individuals who are committed to the given association will be possible in line with own revenues and subsidies. The first-mentioned resources may cover current expenditures to a minor extent; they mostly play a part in the financing of long-term projects. Due to the aforementioned lack of publicity and mechanisms of control and protection, the issuance of bonds or of so-called participation certificates poses difficulties for associations.
We can ultimately state that the registered association will continue to function as one of the most significant legal forms of professional football in Germany, in spite of the fact that the law of associations is far from providing satisfactory conditions for the operations of football clubs carrying out modern business activities. Thus, it is no wonder that in Germany there appears a strong inclination towards the transformation from the form of association into the form of incorporated firm. This does not mean that the clubs (mainly the amateur and mass sports clubs) will turn away from the form of association. Professional football is just such an area. The best example of an efficiently operated association is VfB Stuttgart. 
The traditional form of association as a basic model of management may also be successful in professional football, but it is true that this is realised also thanks to Stuttgart?s main sponsor, its exclusive partners and its team partners. The list is deserving of attention: Gazi, Merzedes Benz, Puma EnBW, Kronbacher, BW Bank, Breuninger, Kärcher, and Coca Cola. Therefore, is no wonder that VfB Stuttgart was and is very successful.

Records of VfB Stuttgart:
German championship:
* Winners (5): 1950, 1952, 1984, 1992, 2007
* Runners-up (4): 1935, 1953, 1979, 2003
German Cup:
* Winners (3): 1954, 1958, 1997
* Runners-up (2): 1986, 2007
German Super Cup:
* Winners (1): 1992
German League Cup:
* Runners-up (3): 1997, 1998, 2005
Oberliga Süd I:
* Winners (3): 1945?46, 1951?52, 1953?54
* Runners-up (3): 1949?50, 1952?53, 1955?56
2nd Bundesliga Süd II:
* Winners (1): 1977
Bezirksliga Württemberg-Baden:
* Winners (2): 1926?27, 1929?30
* Runners-up (1): 1925?26
Gauliga Württemberg:
* Winners (4): 1934?35, 1936?37, 1937?38, 1942?43
* Runners-up (4): 1938?39, 1939?40, 1940?41, 1941?42
UEFA Cup:
* Runners-up (1): 1988?89
UEFA Cup Winners' Cup:
* Runners-up (1): 1997?98
UEFA Intertoto Cup:
* Winners (2): 2000, 2002


Incorporated firms
However, the real concept of an incorporated football firm refers more to the outsourcing of football undertakings and of the fields neighbouring football into incorporated firms, or, respectively, to the transformation thereof into incorporated firms. The surviving association operates as a so-called parent association, and it usually holds the majority of votes in the incorporated football firm.

Share companies
Share companies are advantageous if there is a great need for capital and if it is possible to involve further financers in joint financing. However, in the course of the acquisition of capital, the majority of votes is in each case an indispensable requirement for the parent association, which can be secured through the issuance of registered shares and/or preference shares with restricted transferability (Nagy, 2012a). In the first case, the permission of the management of a share company is also required for the acquisition of shares, while the latter do not secure any votes for the shareholders, and their issuance may not exceed the sum of the shares already issued (ordinary shares and shares with restricted transferability). In this way the formation of an undesirable majority may be prevented; but obviously, the involvement of capital will also be narrowed due to the restrictions, since in the event that a  contribution of further capital is required due to the increase of capital or losses, the parent association may incur obligations of payment in order to secure the majority, because the licence rights are possessed in this case by the parent association. The League Association strictly controls compliance with each regulation. FC Bayern Munich has succeeded in setting up a successful company limited by shares. Successful Football Club: Bayern München Share Company (Figure 3).

Figure 3: Ownership structure


Source: FC Bayern München website

According to pre-agreed plans, up to July 2011, Audi AG increased its share to 9.09% in three phases. As a result, the company approached the existing share of Adidas AG. The association has retained shares making up a quota of 81.2%, while the amount of subscribed capital has risen to EUR 27.5 million. The company limited by shares is the sole owner of the Allianz Arena München Stadion GmbH, which constructed and now operates the stadium. Until 2006, this limited liability company was 50% owned by TSV 1860 München, which then ? due to financial difficulties ? sold its share to FC Bayern München AG for a consideration of EUR 11 million.

Records of Bayern München:
* Most Bundesliga titles won: 21
* Most Bundesliga games won (907) and points achieved (3095)
* Most game days on top of the league: 592
* Most average points per game in the Bundesliga: 1.93
* Most Bundesliga goals scored: 3412
* Most consecutive wins in the Bundesliga (19 March ? 20 September, 2005): 15
* Most games won in a club's first Bundesliga season (1965?66): 20
* Biggest lead over second-place finisher (2002?03): 16 points
* Championship with fewest points under the 3-point rule (2000?01): 63
* Championship with the most losses in a season (2000?01): 9
* As a negative record, Bayern's match in Dortmund in the 2000?01 season was the most unfair match in Bundesliga history, with 15 cards being shown (10 yellow, 1 yellow-red, 2 red), and of those, 12 (8, 1, 1) were shown to Bayern players, which is also a record in Bundesliga history.
* Most championships won: 22
* Most cups won: 15
* Most league cups won: 6
* Most doubles won: 8
* Only club to win the double (Bundesliga and cup) twice in a row (2005?06 and 2006?07)
* Only club to win a championship in both men's and women's football 
* Fastest goal in Champions League history: After 10 seconds by Roy Makaay on 7 March, 2007, against Real Madrid
* Last club to win the Champions League/European Cup three times in a row: 1974?76
* Highest aggregate win in the UEFA Champions League knockout stage: 12?1 on 24 February, 2009, (5?0) and 11 March, 2009, (7?1) against Sporting CP


Limited liability companies
The legal form of limited liability companies is also highly popular in other sports, such as in the field of handball and ice-hockey. Limited liability companies tend to take various forms in football. 
a) One of these forms represents a limited liability football company where all the businesses associated with football are included, and which is generally operated as a subsidiary of a parent company otherwise not involved in football. The license is held by the parent company. In this case, it is the majority owner, the parent association, which determines the foundation of the limited liability company. This solution in fact means the establishment of a subsidiary limited liability company. An example here could be Borussia Mönchengladbach.
b)  The other form of limited liability companies is the entity holding the right to participate in championships, where management by the parent company is also in place, but in this case the entire license comes into the possession of the subsidiary. Nevertheless, all the other areas (marketing, merchandising, etc.) remain with the parent company, or are outsourced to further entities. An advantage of this limited liability company form is that the license has to be secured by the company that uses it, and therefore it is not the entire parent company that needs to comply with the strict licensing requirements, as on the whole the parent company may be in a much worse situation than the limited liability company focusing solely on professional football.
c) The third situation is when the limited liability company plays an essential role within the group, and agrees to apply a broad range of tools for capital acquisition, while preventing external investors from having any significant say in terms of management. 
From among these various solutions, in Hungary it would be expedient to frame a limited liability company representing a real pool of capital resources that is a part of a business group which facilitates access to capital without causing management problems. In this case, the group in fact operates a limited partnership via the limited liability company and where any investor can enter into the position of the limited partner without causing any change in management. Towards this end, Hungary should implement the necessary legal changes, and in addition the attitudes and management practices of clubs also need to be modified.

Limited Partnership with Restricted Liability of Partners
This is a company which has a legal personality, in which, as is well-known, at least one of the partners (the general partner) must bear unlimited and joint and several liability vis-?-vis the creditors. The other owners, who have other interest in the registered capital embodied by the share capital, need not warrant for the liabilities of the company (dormant partners). This is an intermediate legal form that displays the characteristics of a share company and the personal liability existing in the case of limited partnerships, where the general partner provides for the management of the undertaking. The supervisory board elected by the general meeting of the dormant partners supervises the work of the general partner, acting with personal liability, on the one hand, and executes the resolutions of the general meeting, on the other, which includes the dormant partners? declaration of acceptance and agreement in the case of more important decisions. Within the scope of legal forms, such solutions are interesting for football clubs, since they are vested with the characteristics of restricted liability covering the general partner of a limited partnership, and at the same time they open up a lot of alternative opportunities for finding equity. This limited partnership with restricted shareholders? liability also exists in the case of the forms of GmbH & Co. KG.

Figure 4: Organizational Scheme ?Limited Partnerships with share responsibilities?


Source: Dworak, 2010

In the case of the legal form operating with the limited liability of a limited partnership, the parent company as a general partner may not restrict its personal liability. Thus the association should act simultaneously as both general partner and dormant partner if it wishes to retain the aforementioned limited partnership directly as its own property, which is not at all permitted from the legal aspect. Therefore, it is required to insert a limited liability company between the existing association and the limited partnership. This subsidiary LLC is a personally answerable general partner, and it functions as managing director at the same time. The shareholders, as dormant partners, may consist of the parent company itself and of external investors. This has the advantage, on the one hand, that the parent company is secured by an LLC of limited liability, and on the other, that the possibility exists to increase the dormant partners? share in the limited partnership up to almost 100% without violating the rules of the League Association relating to majority share. In Germany, 10 GmbH & Co. KG companies were operating in 2009 in the Licence Leagues (League 1 and 2). See Figure 2.
Should a football GmbH & Co. KG be floated on the stock exchange, it must tolerate a significant handicap as opposed to the companies operating in the form of a share company, if it has also issued preference shares without voting rights. It is a further problem that the possibility of efficiently controlling the leading bodies of an incorporated firm is also missing in the form of football GmbH & Co. KG for potential investors; they must satisfy themselves with the role of dormant partner if they are not members of the parent association. Unfortunately, this may repel potential investors or, at least may be taken into account as a significant negative factor upon the valuation of the participation. It is very important in this solution that separation between the rights of general partners and dormant partners should be complied with strictly in practice. It frequently occurs that potential investors link their possible commitments to being involved in the direction and operation of the undertaking, which will of course appear understandable from the point of view of caring for the safety of their investment. In the case of a football GmbH & Co. KG, any influence by investors may only be implemented within certain limits. By the way, this is shown by the objection made by the League Association against an investor agreement entered into by TSV 1860 München in the 2009 spring season. The investor agreement contained decisions in respect of persons as conditions for contract, which would have had the consequence of illicit influence on the side of the dormant partner, and/or it would have secured the right of veto in respect of the decisions of the associations.
We can finally state that the outsourcing of significant branches of a football undertaking into the form of an incorporated firm should be deemed a considerable step forward in comparison with the deficient regulations of the operations of associations, regarding the economic professionality of football clubs, partially from the side of commercial law and partially from the side of shareholding law. Through the introduction of the organizational and legal structures necessary for football clubs to be considered professional, there are further positive effects generated for football companies: for example, the football undertaking has to reevaluate its assets due to the transformation, which usually has a positive effect on the proportion of equity. This is still complemented by the fact that, in the event of incorporation, equity will increase through the accumulation of profits based on the compulsory regulations of the application of profits. Incorporation allows external investors to get involved in football undertakings, which may be implemented through investment into the respective business quotas of an LLC or a share company or a limited partnership. Here we must stress above all the obligation of publicity and information, which makes the incorporated firms transparent for potential investors of debt capital, and allows for reliable credit ratings and risk assessments. Furthermore, it is also advantageous that incorporated football firms operate similarly to incorporated firms functioning in other fields, from the point of view of the protection of creditors. Finally, a large palette of alternative financing is available for incorporated firms which are not available for the original legal form and organizational structure of associations. Nevertheless, it should be underlined that, in the event of the use of the form of the incorporated football firm, it is about the transformation and change of the organizational unit of execution and not the total abandonment of the form of association. The form of association will continue to be the main form for football undertakings and a fundamental solution for the procurement of capital. Due to the reasons above, in this respect professional football is partially exempt.
IV. Changed business model
As a result of the above changes, the business model of professional football has been fundamentally transformed.

Table 2. Comparison of the two models of football
Source of financing government revenuesbusiness revenuesRole of budget constraintssoft budget constraintshard budget constraintsFeature of the ownerno real ownerreal owner(s)Operational frameworkNon-profit:
form of social clubcorporate form of companyRole of footballlegitimacy of the political system,
improving moraleservice as a subfield of the entertainment industrySource: András, 2004

In the case of developed football countries and market economies, the business model has clearly prevailed in professional football. While in the developing countries, in spite of all the required structural changes, a mixed system has been formed that still bears the elements of the state model.

V. Organizational requirements for professional football entities in Switzerland
In Switzerland, the Football League stipulates rigorous reporting obligations, and in addition it prescribes the structures and mandatory statuses for the operation of a Super League team. Relying on these foundations, right-minded Calvinistic Puritanism operates all the clubs in highly structured forms. Therefore, I would also recommend the introduction of the Swiss organizational requirements in OTP Bank League. 

Swiss mandatory minimum organizational requirements for the teams of the Super League:
1 person, UEFA-licensed trainer
1 person, assistant trainer with Super League license
1 person, goalkeeper trainer with Super League license
1 person, fitness trainer
1 person, managing director and administration manager (the same person can work in both functions)
1 person, finance manager
1 person, marketing manager
1 person, media manager
Recommended: 1 person, security expert, possible cooperation with the representatives of supporters (Lehmkuhl and Siegrist, 2009).
In Switzerland, any omission of monthly reports or non-fulfilment of payment obligations would promptly entail a fine of CHF 10,000. In response to further omissions, negotiations are commenced in relation to the withdrawal of the license.
VI. Teams of the OTP Bank League by legal forms of business and organizational forms
At first sight, it can be claimed that there is nothing wrong with clubs participating in the OTP Bank League in terms of their restructuring into capital companies, as in Hungary the formal conversion of first-league clubs into capital companies has actually been implemented (Figure 5). In Switzerland, this restructuring move has been executed on the basis of the Association's mandatory requirement. In contrast, as has been pointed out above, the German first league still has a number of clubs operating in the form of associations. Restructuring into capital companies has also been implemented in the Greek first league on a mandatory basis.


Figure 5. Teams of OTP Bank League by legal forms of business and organization

OTP Bank LeagueFootball ClubLegal FormGyõri ETOlimited liability companyVIDEOTONlimited liability companyBudapesti Honvédlimited liability companyMTK Budapestshare companyFerencvárosi TCshare companyDVSC-TEVAshare companyKTE-Phoenixlimited liability companyÚjpest FClimited liability companyDVTKlimited liability companyKaposvári Rákóczilimited liability companyPMFC Matiaslimited liability companyMVM-Pakslimited liability companyLombard Pápa Termál FClimited liability companyFGSZ Siófoklimited liability company
Source: Own research

Nonetheless, another issue is how much in Hungary the preferred form of capital companies ensures additional capital resources (Nagy, 2012b). For instance, a club achieving successes both in the sport and economic fields currently operates as a private company limited by shares, with a single private person being its sole owner. Here, the question is how much the form of a one-man company serves the ends of capital acquisition, as the amount of the capital subscribed by the sole owner is only HUF 49 million. On the other hand, one can find a limited liability company with two legal entity partners and registered capital of more than one billion Hungarian forints. The individual situations can be highly varied.  In view of the unfortunate, yet typical lack of capital resources that is a dominant feature of domestic football in spite of the corporate income tax allowances and preferential aids, I would propose the establishment of real capital companies, especially the operation of groups described above (Borussia Dortmund). Similarly, in Hungarian football, such capital companies have to be established that are able to attract additional, fresh capital resources. A large proportion of business processes ? primarily in the field of ticket sales and player trading transactions ? remain outside the balance sheets. Considerable problems are caused by kickbacks provided on corporate income tax allowances and other aids, as well as other uncertain and invisible incomes and expenses (Kotlán, Machová and Janíèková, 2011).
In several respects, the government would be interested in elevating 3-4 clubs in Hungary to make them capable of outstanding and high-standard sporting and economic performance, which would have economic, budgetary benefits ? not to mention the social implications that cannot be directly assessed by means of economic indicators (Bács and Szima, 2012).
The balance sheet figures of DVSC Futballszervezõ Zrt. reflect significantly positive results for the past four years, and, moreover, the owners were paid dividends in an amount of HUF 800 million based on 2009 profits, while a decision was made on the establishment of the training center of the Debrecen Football Academy, an investment of HUF 800 million (Bács and Szilágyi, 2012).
VII. Establishment of controlling organizational units at football enterprises 
In Germany, the teams of Leagues 1?3 were involved in a questionnaire survey concerning the application of controlling. 53 associations and capital companies responded to the questionnaires, which indicated and extremely positive approach to the effort (Table 3).

Table 3. Application of controlling at German football clubs
Time of application of controllingDistribution (%)1-5 years58,55-10 years13,7more than 10 years17,1no answer10,3Source: Controlling im Profifußball, Roßbach, Th:2011, p. 43

It is an even more interesting question to see what means of controlling the clubs of Leagues 1?3 use in Germany ? some of the clubs apply more than one of these tools (Table 4).

Table 4. Applied controlling devices and methods in Leagues 1?3. of Germany
Controlling devices and methodsRate of application (%)Benchmarking27.27Comparison of planned and factual figures12.12Balanced Scorecard9.09Obtainment of liquidity6.06Rolling forecasts3.03Scenario procedures3.03Profit and loss analysis3.03Detailed budgeting3.03Target cost analysis3.03Quarterly and monthly reports3.03Addison controlling tool3.03Own analyses3.03Deviation analyses3.03ABC analysis3.03Business plan3.03Others3.03Source: Controlling im Profifußball, Roßbach, Th:2011, p.  44

An important question is what are the controlling targets, i.e. what is the focal point of control in football enterprises (Table 5)

Table 5. The subject-matter of controlling in Leagues 1?2 of Germany
Subject of controllingIntensity (%)merchandizing 79.3payments89.7operations 89.7stadium use 79.3media rights 82.8sponsorship96.6player transfers, 69.0marketing89.7Source: Controlling im Profifußball, Roßbach, Th:2011, p. 46

Considering that in Hungary many first-league clubs prepare only annual budgets to serve as the basis of economic management, one could claim that it is not only goal scoring and excellent defense where Hungary has gaps to close nowadays, but also in the field of the application of developed methods of a market economy. Long-term successes can be achieved only with a firmly grounded, stable organizational and financial background (Borbély, 2008, Nagy, 2013).
VIII. Conclusion
In summary, it has to be pointed out that financial foresight and controlling constitute such indispensable subsystems for football associations and enterprises responsible for the alignment and coordination of various other processes and subsystems. With reliance on the set of indicators of controlling, a higher level of transparency is put in place to support the supply of information and managerial decisions. As a result, environmental changes become recognizable in their early stages, and therefore the necessary measures can be taken sooner, which then allows active management. Furthermore, controlling promotes the continuous review and correction of own activities. For this reason, it is very much expedient to set up a planned, documented and consciously implemented system of controlling both in the operative and strategic fields. The above-described tools that have already been partly applied by the German clubs ? such as the Balanced Scorecard ? are eminently suitable for helping to find adequate responses to future challenges. These means of controlling can be combined with each other, and therefore the Balanced Scorecard system can be structured together with a risk management system and detailed budgeting (Haas, 2006).
Controlling has an increasingly important role in the operation of foreign football clubs, though its potential still has not been exploited optimally in major football nations. The processes will be accelerated by UEFA's new licensing regulations posing strict requirements and attributing an even more significant role to financial stability (financial fair play). Football is becoming increasingly professional and business-oriented, a process which requires the application of modern management systems. The application of controlling is spreading, which improves the effectiveness of the business areas. This relatively new, gradually strengthening aspect, i.e. the economic side of football, seems to be extremely important for the clubs and the federations alike via the issuance of licenses. Additional stakeholders include football-related market vendors, agents, owners, as well as shareholding companies, creditors and even auditors. Abroad, in the major football nations, the current focal point of debate is the organizational forms of controlling systems. Experts deem it to be expedient to operate independent controlling departments organized in the light of the individual controlling tasks at the football clubs.

References
Allen, F. (2011). Liquidity and crises. Oxford: Oxford University Press.
András, K. (2004). A hivatalos labdarúgás piacai. 53.sz Mûhelytanulmány. Budapesti Közgazdaságtudományi és Államigazgatási Egyetem, 2004.
Bács, Z., Szima G. (2012). Értékek a klubépítésben (és ami még mögötte van). Nemzetközi Turizmus és Sportmanagement Konferencia, Debrecen, 1-12.

Bács, Z., Szilágyi S. (2012). A DVSC BL szereplésének gazdasági és államháztartási vonatkozásai.  Debreceni Egyetem, 1-6.
Borbély, A. (2012). Országos Sporttudományi Kongresszus. Magyar Sporttudományi Szemle, 12(2), 6-12.
Borbély, A. (2008). Sportpolitika és sportgazdaság változásai Magyarországon és az Észak- alföldi Régióban, különös tekintettel a küzdõsportokra. (habilitációs anyag)
Bott, S. (2007). Internationalisierungsstrategien von nationalen Fußballigen. 1. Auflage, Books on Demand GmbH, Norderstedt.
Dworak, A. (2010). Finanzierung für Fußballunternehmen. Berlin: Erich Schmidt Verlag.
Haas, O. (2006). Controlling der Fußballsunternehmen. Auflage 2. Berlin: Erich Schmidt Verlag.
Keller, Ch. (2006). Corporate Finance im Profifussball. Stuttgart: Ibidem Verlag.
Kotlán, I., Machová, Z., Janíèková, L. (2011). Taxation Influence On The Economic Growth. Politická ekonomie, 59(5), 638?658.
Lehmkuhl, D., Siegrist, O. (2009). Swiss Football. Retrieved form: http://www.ipw.unisg.ch/org/ipw/web.nsf/SysWebRessources/Presentation+Lehmkuhl+Swiss+Football/$FILE/SwissFootball.pdf
Nagy, I. Z. (2011). The Fight against Income Evasion in Hungary. Acta Polytechnica Hungarica, Journal of Applied Sciences, 8(2), 103-116.
Nagy, I. Z. (2012a[P2]). Modern Forms of Business in Professional Football. DANUBE: Law and Economics Review, 3(2), 1-17.
Nagy, I. Z. (2012b). Die Rolle der Rechtsformen in der Führung des professionelen Fußballs. In proccedings of the 10th International Conference on Management, Enterprise and Benchmarking (MEB), Budapest, 2012, 193-209.
Nagy, I. Z. (2013). A sikeres vállalkozás pénzgazdálkodása és pénzügyi tervezése. In Vállalkozásfejlesztés a XXI. században III.: tanulmánykötet, Budapest: Óbudai Egyetem, 2013, 7-34.
Nagy I. Z., Való, V. (2013). The Lawful Execution of Employer Rights Is an Important Subsystem of Corporate Governance (A Labour Case Study). DANUBE: Law and Economics Review, 4(3), 201?217.
Roßbach, Th. (2011). Controlling im Profifußball. BiTS Business and Information Technology School GmbH. (Diplomarbeit) Iserlohn, Németország.
1 University of Debrecen, Faculty of Applied Economics and Rural Development, Böszörményi u. 138, H-4032 Debrecen, Hungary. E-mail: babae@agr.unideb.hu.
[P1]Pouze tato jedna autorka, bez Nagyho
[P2]Rozli¹it 2012a a 2012b
---------------

------------------------------------------------------------

---------------

------------------------------------------------------------

